---
title: "Practical Machine Learning Course Project"
author: "Rajat Gupta"
date: "September 3, 2017"
output:
  word_document: default
  html_document: default
---

## 1. Overview

This document is the final report of the Peer Assessment project from Coursera's course Practical Machine Learning, as part of the Specialization in Data Science. It was built in RStudio, using its knitr functions, meant to be published in html format. This analysis meant to be the basis for the course quiz and a prediction assignment writeup. The main goal of the project is to predict the manner in which 6 participants performed some exercise as described below. This is the "classe" variable in the training set. The machine learning algorithm described here is applied to the 20 test cases available in the test data and the predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.

## 2.Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##3. Data Loading and Exploratory Analysis

### 3a. Data Source
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.


### 3b. Environment Setup
```{r library, echo=TRUE}
library(caret)
library(randomForest) #Random forest for classification and regression
library(rpart) # Regressive Partitioning and Regression trees
library(rpart.plot) # Decision Tree plot
# setting the overall seed for reproduceability
set.seed(9999)

```

### 3c. Loading data sets and preliminary cleaning

First we will load the data sets in R and remove the missing values.
Results will be hidden from the report for clarity and space considerations.

```{r cleaning}
# We notice that both data sets contain columns with all missing values - these will be deleted. Some missing values are coded as "#DIV/0!" or "" or "NA" - these will be changed to NA.

# Loading the training data set into my R session replacing all missing with "NA"
testing <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))

# Loading the testing data set 
training <- read.csv('pml-training.csv', na.strings=c("NA","#DIV/0!", ""))

# Check dimensions for number of variables and number of observations
dim(training)
dim(testing)

# Delete columns with all missing values
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]

# Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]

```

### 3d. Partitioning the training data set to allow cross-validation

The training set is partionned into 2 sets: subTraining (75%) and subTest (25%).

```{r partintion, echo=TRUE}
set.seed(9999)
subsamples <- createDataPartition(y=training$classe, p=0.75, list=FALSE)
subTraining <- training[subsamples, ] 
subTesting <- training[-subsamples, ]
```

### 3e. A look at the Data

The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.

```{r chart, echo=TRUE}
plot(subTraining$classe, col="red", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="Classe levels", ylab="Frequency")
```

## 4 Prediction models
### 4.a Decision Tree

```{r DT, echo=TRUE}
set.seed(9999)
model1 <- rpart(classe ~ ., data=subTraining, method="class")

# Plot of the Decision Tree
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)

# Prediction and test results in subTesting data set:
prediction1 <- predict(model1, subTesting, type = "class")

confusionMatrix(prediction1, subTesting$classe)
```

### 4.b Using Random Forest

```{r RF, echo=TRUE}
set.seed(9999)
model2 <- randomForest(classe ~. , data=subTraining, method="class")

#Prediction and test results in subTesting data set:
prediction2 <- predict(model2, subTesting, type = "class")

confusionMatrix(prediction2, subTesting$classe)

```

##Decision

As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to ~ 0.74 (95% CI: (0.73, 0.75)) for Decision Tree model. The random Forest model is choosen. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

## Submssion

```{r Sub, echo=TRUE}

# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model2, testing, type="class")
predictfinal

# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)

```

 Thank you for your time.